On The Impact of Atoms of Confusion in JavaScript Code

### 2021-10-21 2_Background.tex
Passed through Sections 1 and 2. Stopped at the beginning of Page 4. I will resume tomorrow.

The concept of program comprehension is central to software maintenance and feature development \cite{DBLP:conf/iwpc/TilleySP96, DBLP:journals/ibmsj/OHareT94}. Atoms of confusion is an important field of program comprehension introduced by Gopstein et al. \cite{DBLP:conf/msr/GopsteinZFC18}, who defined them as small code patterns that can verifiably lead to misunderstandings. In addition, for a code pattern to be considered an atom, there must be some alternative pattern or language construct that is functionally equivalent and less likely to cause confusion. One example that occurs in many programming languages is the Change of Literal Encoding. For instance, the \clang code statement \lstinline{printf("\%d", 013);} 
often leads programmers to predict the output to be \texttt{13}, even though the correct answer is \texttt{11}. This occurs because a leading \texttt{0} in a numerical literal indicates that the number is in base \texttt{8}, a fact that is not only unknown to less experienced programmers, but misleading even for seasoned developers. After formulating a list of 19 atom candidates, Gopstein et al. \cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} were able to identify 15 code idioms and language constructs that present a statistically significant difference in comparative answer correctness when each of the 15 atoms was removed. The least confusing atom showed a 14\% boost in prediction accuracy when the atom was removed, whereas the most confusing one showed a 60\% accuracy increase. In a different work, Gopstein et al. \cite{DBLP:conf/msr/GopsteinZFC18} presented the results of a comprehensive research on the incidence of atoms of confusion in the wild, considering open-source projects written in \clang and \cpplang~programming languages.

@@

These previous studies~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17,DBLP:conf/msr/GopsteinZFC18}
have a great influence in our work. Even though JavaScript is a dynamic language that differs from \clang in a variety of aspects, it also has a number of constructs in common. In addition, the imperative aspects of the language are syntactically similar to \clang, e.g., assignments, conditional expressions, \texttt{if}-statements, pre and post increments and decrements, among others. On the one hand, this means that some of the atoms of confusion that exist in \clang programs may also occur in JavaScript code. On the other hand, differences between the languages and the surrounding programming cultures may lead to code patterns that are atoms of confusion in one language not being atoms in the other one.
%Although JavaScript allows us to program in different paradigms from the one C follows, most of the JavaScript code found in popular repositories are written under the imperative paradigm. Also, both languages share many syntactical rules, which allows us to replicate some of the C atoms in JavaScript.

@@

%Gopstein et al. also influenced researchers to look into atoms in languages other than \clang and \cpplang.
Langhout and Aniche~\cite{Langhout:2021:ACJ} derived a set of 14 atoms for Java and performed an experiment with 132 students. They found that participants are more likely to make mistakes (ranging from 2.7 up to 56 times) when atoms are present in half of their studied atoms. Castor~\cite{castor2018} presented a preliminary catalog of six atom candidates for the Swift programming language. Unlike JavaScript, in Swift, most of the atoms identified by Gopstein et al.~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} are avoided by construction, e.g., it does not have assignments as values, increment operators, or macros. 
%The paper also presents a structured definition for what it means for a code pattern to be an atom of confusion and a set of general principles to help in the identification of new atoms.

@@

Santos and Gerosa~\cite{Santos:2018:ICP} analyze the impact of eleven coding practices on program readability. However, most of the practices analyzed in their study relate to code formatting, instead of language constructs or idioms that might introduce misunderstanding. We borrow from their research the methodology for collecting the practitioners opinion about code snippets preference.

>>>>

The concept of program comprehension is central to software maintenance and feature development \cite{DBLP:conf/iwpc/TilleySP96, DBLP:journals/ibmsj/OHareT94}. The study of program comprehension in the presence of atoms of confusion was introduced by Gopstein et al. \cite{DBLP:conf/msr/GopsteinZFC18}, who defined them as small code patterns that can verifiably lead to misunderstandings. In addition, for a code pattern to be considered an atom, there must be some alternative pattern or language construct that is functionally equivalent and less likely to cause confusion. One example that occurs in many programming languages is the Change of Literal Encoding. For instance, the \clang code statement \lstinline{printf("\%d", 013);} 
often leads programmers to predict the output to be \texttt{13}, even though the correct answer is \texttt{11}. This occurs because a leading \texttt{0} in a numerical literal indicates that the number is in base \texttt{8}, a fact that is not only unknown to less experienced programmers, but misleading even for seasoned developers. After formulating a list of 19 atom candidates, Gopstein et al. \cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} were able to identify 15 code idioms and language constructs that present a statistically significant difference in comparative answer correctness when each of the 15 atoms was removed. The least confusing atom produced a 14\% boost in prediction accuracy when it was removed, whereas the most confusing one showed a 60\% accuracy increase. In a different work, Gopstein et al. \cite{DBLP:conf/msr/GopsteinZFC18} presented the results of a comprehensive study on the incidence of atoms of confusion in the wild, considering open-source projects written in \clang and \cpplang~programming languages.

@@

These previous studies~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17,DBLP:conf/msr/GopsteinZFC18}
have a great influence in our work. Even though JavaScript is a dynamic language that differs from \clang in a variety of aspects, it also has a number of constructs in common. In addition, the imperative aspects of the language are syntactically similar to \clang, e.g., assignments, conditional expressions, \texttt{if}-statements, pre and post increments and decrements, among others. On the one hand, this means that some of the atoms of confusion that exist in \clang programs may also occur in JavaScript code. On the other hand, differences between the languages and the surrounding programming cultures may lead to code patterns that are atoms of confusion in one language not being atoms in the other one.\castor{Shouldn't we explain here how this work differs from previous work on the C language \textbf{besides} just saying that JavaScript is different? I think we should at least emphasize the different experimental setup. I think this is very important to emphasize how different this study is from the previous ones. }
%Although JavaScript allows us to program in different paradigms from the one C follows, most of the JavaScript code found in popular repositories are written under the imperative paradigm. Also, both languages share many syntactical rules, which allows us to replicate some of the C atoms in JavaScript.

@@

%Gopstein et al. also influenced researchers to look into atoms in languages other than \clang and \cpplang.
Langhout and Aniche~\cite{Langhout:2021:ACJ} derived a set of 14 atoms for Java and performed an experiment with 132 students. They found that participants are more likely to make mistakes (ranging from 2.7 up to 56 times) when atoms are present in half of their studied atoms\castor{I do not understand the meaning of the previous section "atoms are present in half of their studied atoms". Furthermore, I do not understand the factors between parentheses.}. Castor~\cite{castor2018} presented a preliminary catalog of six atom candidates for the Swift programming language. Unlike JavaScript, in Swift, most of the atoms identified by Gopstein et al.~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} are avoided by construction, e.g., it does not have assignments as values, increment operators, or macros. 
%The paper also presents a structured definition for what it means for a code pattern to be an atom of confusion and a set of general principles to help in the identification of new atoms.

@@

Santos and Gerosa~\cite{Santos:2018:ICP} analyze the impact of eleven coding practices on program readability. However, most of the practices analyzed in their study relate to code formatting, instead of language constructs or idioms that might introduce misunderstanding. In addition, the study participants did not have to perform comprehension tasks. We borrow from their research the methodology for collecting the practitioners opinion about code snippet preferences.\castor{I would rather we did not say this about our study because this methodology of presenting two or more code snippets and asking subjects which one they prefer is fairly old, e.g., the following paper already used this approach: Arab M (1992) Enhancing Program Comprehension : Formatting and Documenting.ACM SIGPLAN Notices 27(2) :37–46}.

### 2021-10-21 3_Method.tex
Passed through Sections 1 and 2. Stopped at the beginning of Page 4. I will resume tomorrow.

Answering these research questions has several implications. For instance, we can either generalize or refute the perceptions about atoms of confusion already discussed in the literature (goal of research questions (RQ1) and (RQ2)). In addition, answers to the third and fourth questions allow us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.
We conduct a mixed-methods study to anwer these questions, including a survey, a set of interviews, and a study for mining JavaScript software repositories.

@@

In the first study, we explore the impact of atom candidates on
understanding JavaScript code (goal of research question RQ1).	
To this end, we investigate whether or not code snippets that
contain atom candidates (a) produce a higher
misunderstanding rate when programmers
try to predict their outcome and (b) require
programmers to take longer to predict their output.
Since JavaScript and \clang have some constructs in common (Section~\ref{back}), we first selected a set of 9+1 atom candidates: 9 atoms of confusion already validated~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} for the \clang language that also exist in JavaScript programs + 1 atom candidate that is specific of the JavaScript language (Automatic Semicolon Insertion).
Our supplementary material discusses and shows examples of the atoms we consider in our research.

@@

The design of our first study blocks two variables
(developer experience and the code snippets) and
considers two treatments: the presence or absence of 
atoms candidates within the code snippets. 
To achieve such a design goal, controlling the effect of experience and individual code snippets, we resorted to the \textit{Latin Square Design} \cite{Hunter-Experimenters}. Using this design we create a 2 x 2 matrix in which each row represents a subject and each column indicates the set of code snippets. The design of each square (a replica) is such that no treatment is repeated in the same row or column. For example, considering that we have a set of 10
code samples $s_0, s_1, ..., s_{10}$, if a given subject (P1) is asked to predict the output of the code samples $s_0, s_1, ..., s_5$ that contain atom candidates, then, when answering questions about non-confusing code snippets, they will only be presented with non-confusing versions of the code samples $s_6, s_7,..., s_{10}$. Furthermore, a given subject (P2), which constitutes the second row of our example square, will be asked questions about the non-confusing versions for $s_1, s_2, ..., s_5$, and will answer questions about confusing snippets for $s_6, s_7,..., s_{10}$. By doing that, we guarantee that all versions of the code snippets are contained within each square, and that each configuration occurs only once within a square. Figure \ref{fig:latinsquare} offers a visual representation of the concept.

>>>>

Answering these research questions has several implications. For instance, we can either generalize or refute the perceptions about atoms of confusion already discussed in the literature (goal of research questions (RQ1) and (RQ2)). In addition, answers to the third and fourth questions allow us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.
We conduct a mixed-methods study to answer these questions, including a survey, a set of interviews, and a study for mining JavaScript software repositories.

@@

In the first study, we explore the impact of atom candidates on
understanding JavaScript code (goal of research question RQ1).	
To this end, we investigate whether or not code snippets that
contain atom candidates (a) produce a higher
misunderstanding rate when programmers
try to predict their outcome and (b) require more time for programmers to predict their output.
Since JavaScript and \clang have some constructs in common (Section~\ref{back}), we first selected a set of 9+1 atom candidates: nine previously-validated atoms of confusion for the \clang language~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} that also exist in JavaScript programs plus one atom candidate that is specific to the JavaScript language (Automatic Semicolon Insertion).
Our supplementary material discusses and shows examples of the atoms we consider in our research.

@@

The design of our first study blocks two variables
(developer experience and the code snippets) and
considers two treatments: the presence or absence of 
atoms candidates within the code snippets. 
To achieve such a design goal, controlling the effect of experience and individual code snippets, we resorted to the \textit{Latin Square Design} \cite{Hunter-Experimenters}. Using this design we create a 2 x 2 matrix in which each row represents a subject and each column indicates the set of code snippets. The design of each square (a replica) is such that no treatment is repeated in the same row or column. For example, considering that we have a set of 10
code samples $s_0, s_1, ..., s_{10}$, if a given subject (P1) is asked to predict the output of the code samples $s_0, s_1, ..., s_5$ that contain atom candidates, then, when answering questions about non-confusing code snippets, they will only be presented with non-confusing versions of the code samples $s_6, s_7,..., s_{10}$. Furthermore, a given subject (P2), who constitutes the second row of our example square, will be asked questions about the non-confusing versions for $s_1, s_2, ..., s_5$, and will answer questions about confusing snippets for $s_6, s_7,..., s_{10}$. By doing that, we guarantee that all versions of the code snippets are contained within each square, and that each configuration occurs only once within a square. Figure \ref{fig:latinsquare} offers a visual representation of the concept.
